Simulating physiological flexibility in the acute glucocorticoid response to stressors reveals limitations of current empirical approaches

### 2022-05-05 3_r_scripts/Manuscript.Rmd
Prepping for revision.

---
title: "Functions for simulating data and designing studies of physiological flexibility in the acute glucocorticoid response to stressors"
author: 
 - Conor C. Taff
header-includes:
  - \usepackage[font={footnotesize}, labelfont={bf}]{caption}
fontsize: 12pt
output: 
  bookdown::word_document2:
      number_sections: FALSE
      toc: FALSE
      extra_dependcies: ["flafter"]
  bookdown::pdf_document2:
      number_sections: FALSE
      toc: FALSE
      extra_dependencies: ["flafter"]
bibliography: references.bib
csl: animal-behaviour.csl
---

@@

@@

Wild animals often experience unpredictable challenges that demand rapid and flexible responses. The glucocorticoid mediated stress response is one of the major systems that allows vertebrates to rapidly adjust their physiology and behavior. Given it's role in responding to challenges, evolutionary physiologists have focused on the consequences of between-individual and, more recently, within-individual variation in the acute glucocorticoid response. Although sophisticated approaches are available to partition this variation statistically, empirical studies of physiological flexibility are severely limited by the logistical challenges of measuring the same animal multiple times during a single acute response or across multiple instances of acute responses. Empiricists have largely adopted the strategy of standardizing sampling as much as possible to allow for comparison between individuals, but this standardization also makes it very difficult to detect certain types of variation in the functional shape of acute response curves. Data simulation is a powerful approach when empirical data are limited, but has not been adopted to date in studies of physiological flexibility. In this paper, I describe the `simcoRt` package, which includes functions that can generate realistic acute glucocorticoid response data with user specified characteristics. Simulated animals can be sampled continuously through an acute response and across as many separate responses as desired, while varying key parameters (e.g., the degree of correlation between the speed and scope of a response). Using this simulation, I explore several possible scenarios to highlight areas where simulation might either provide new insight into physiological flexibility directly or aid in designing empirical studies that are better able to test the hypotheses of interest. All of the code required to reproduce the examples presented here is included in a public GitHub repository.

@@

The recognition that characterizing the functional shape of an acute stress response is challenging goes back to the earliest studies conducted in wild animals. Early studies often employed various control groups and sampled individual animals at a variety of time points over a long period in order to describe the full response curve for a particular group [e.g., a species or a breeding stage, @wingfield1992]. These validations were considered essential to characterize key parameters of the acute response for each group being studied (i.e., baseline, rate of increase, maximum level, time of peak, and area under the curve; John Wingfield, personal communication). The challenge of estimating these parameters becomes much more difficult when trying to describe the response for an individual animal rather than for a group, because glucocorticoids can often only be measured at two or three time points and only a small number of times per animal [e.g., @vitousek2018]. Because these studies require an estimate for each individual, the solutions used by older studies that added additional animals to allow for sampling at more time points are not available.

@@

One of the characteristics of both the within-individual reaction norm and FVT literature is that empirical work has proceeded in very close coordination with simulation and statistical method development. In contrast, studies of endocrine flexibility often point to these methods, but don't address the ways that the particular logistical challenges of hormone measurement might suggest altered study designs. While many of the tools developed in these fields are transferable, studies of physiological flexibility would benefit from a focus on analysis development and testing that explicitly incorporates the particular details of these questions. One way to accomplish these goals is to use simulations, but to my knowledge no studies of physiological flexibility have developed simulations of the acute stress response that address the issues discussed above.

@@

## DESCRIPTION OF THE SIMULATION

I developed a package called `simcoRt` in R version 4.0.2 [@rcore] with functions to generate acute physiological response curves. This simulation makes no assumptions about the  mechanistic process that results in the shape of a glucocorticoid response. Rather, parameters are sampled to generate curves that are similar in shape and degree of variation to empirically observed responses (Figure \@ref(fig:concept-fig)). This simulation is designed to create data sets with realistic structure that can be used to better design and plan studies of physiological flexibility, to evaluate power of current study designs, and to evaluate the sensitivity of sampling regimes to any number of modifications to the shape of glucocorticoid response curves (e.g., changing covariation patterns between different features of the response). I explore a small number of scenarios in the next section, but I expect that many other scenarios can be addressed with these tools. For illustration purposes, I refer to simulated glucocorticoid responses, but the simulation applies equally well to any physiological mediator of a rapid response. The package can be installed in R using the following command.

@@

A second function, `cort_sim2`, starts with a population of animals generated from `cort_sim1` and samples observed acute glucocorticoid responses an arbitrary number of times for each animal. Two sources of variation in the observed relative to true parameter values can be specified. First, within-individual variation in expression is represented by specifying what amount of variation in the observation of each parameter is determined by the true value and what amount is determined by an additional randomly sampled response, based on the population parameters (this additional sampling maintains the user specified covariance structure of the population). After sampling the parameters, values are interpolated for each one minute time point and a localized regression is fit to create a smoothed curve that represents the observed glucocorticoid response. From this expressed response, individual data points are then collected at user specified times that would reflect an empirical study design (e.g., 2, 30, and 60 minutes). Additional noise is added to these data points to represent measurement error (e.g., assay error).

@@

Finally, given recent interest in estimating the repeatability of glucocorticoid regulation [@taff2018; @cockrem2013; @hau2016], I also included a function that takes input from `cort_sim2` and calculates the observed repeatability of several measures using package `rptR` [@stoffel2017]. Full details are included in the package documentation, but this function returns repeatability for each individual time point specified in the down sampled data set, profile repeatability [@reed2019], and repeatability for area under the curve calculated as both increase (AUC~I~) and ground (AUC~G~) approaches [@pruessner2003]. For each AUC measure, the function returns repeatability for the full time course, for an estimate using only the observed values in the down sampled data set, and for the full data set constrained to the time period encompassing the observed data points. Simple plots illustrating repeated samples from the same individuals are also returned by default.

## EXAMPLE APPLICATIONS OF SIMULATION

@@

### *Simulating empirically parameterized data*

@@

### *Accurately measuring a single glucocorticoid trait*

Single time point measures of glucocorticoids are often interpreted as representing meaningful variation between individuals. For example, variation in the level of glucocorticoids after 30 minutes of standardized restraint is typically interpreted as variation in the magnitude of the stress response [@taff2019]. However, this interpretation rests on assumptions that are rarely explicitly tested with empirical data. For example, the time chosen to take a stress-induced sample is often assumed to be either at the species peak or during a plateau period after the species peak. In some early studies, great care was taken to determine an average population level peak time [CITE], but many studies adopt the widely used 'standard' time of 30 minutes post capture without extensive validation. While there is a general assumption that sampling later than the peak is acceptable (and perhaps preferable) because animals will be sampled during a relatively stable high plateau, there is little empirical data to evaluate this assertion or to determine how much under or overshooting the species peak timing might influence inferences. Furthermore, even when the average peak timing is well established, differences in the amount of between-individual variation in the time to reach the peak or in peak values are common across species and even in different life history stages within species [CITE]. The combinations of these patterns of variation could have major consequences on the accuracy of single point estimates taken at 30 minutes, but these questions cannot be addressed directly with empirical datasets where the true underlying values of each individual are unknown.

@@

Results of this simulation are summarized in figure \@ref(fig:max-sim). The amount of between-individual variation in the maximum glucocorticoid value has a profound effect on the ability to detect true maximal levels with samples taken at 30 minutes. In one sense, this result is unsurprising because it is intuitive that large differences would be easier to detect, but there are important consequences of this fact for interpreting studies that seek to link between-individual variation in the magnitude of the stress response with other traits. For example, the magnitude of the acute stress response often varies substantially across life history stages [CITE]. Even if study designs are identical and maximum glucocorticoids are associated with performance, it will be easier to detect those patterns during life history stages with greater variation (see section on detecting fitness associations below). There is a weaker, but still substantial impact of variation in the time taken to reach maximum values on the accuracy of estimates in this simulation. Greater variation in time reduces the accuracy of estimates of maximal values. Finally, the timing of sampling relative to the average population peak timing also influences accuracy. Measuring after the average peak time results in the most accurate estimates across a range of parameter values, while measuring before the average peak time produces the least accurate measures, particularly when there is also high variation in the time to reach maximum values between individuals. This simple example demonstrates clearly that the same experimental design will perform better or worse depending on the combination of glucocorticoid regulation parameters in the population being studied.

### *Exploring covariance between response components*

In reality, fully characterizing the acute glucocorticoid response requires more than identifying just the maximum value reached. Individuals may differ in baseline levels, rate of initial increase, the speed of reaching the maximum level, time spent at maximum, and the speed of return to baseline. Moreover, each of these components of the endocrine response could be positively or negatively correlated with each other within and between individuals. In these cases, measurements taken at particular time points contain information about multiple aspects of the response and without additional information it may be difficult to know what trait is being measured. The fact that each of these traits might be important and that they might covary has been discussed in a general sense [CITE], but simulations are uniquely powerful for exploring under exactly what conditions time point measure of glucocorticoids can or cannot be used as indicators of these traits.

@@

To illustrate this point, I explored the consequences of variation in the correlation between and relative amount of variation in just two aspects of the acute stress response: the maximum glucocorticoid level reached and the time required to reach the maximum level. For simplicity, I refer to the 'speed' of the response, but note that other aspects, such as the rate of initial increase, could also be considered as variation in the speed of response. When considering these two traits, a population of animals could plausibly display one of three patterns. Individuals that reach their maximum value faster might also reach higher values (figure \@ref(fig:cor-ex1)A; simulation correlation = -0.6). Alternatively, the speed and maximum values might vary independently (figure \@ref(fig:cor-ex1)B; correlation = 0). Finally, individuals that are faster might max out at lower glucocorticoid values (figure \@ref(fig:cor-ex1)C; correlation = 0.6). While many researchers in this field might have intuitions about which of these scenarios is most likely to prevail, there is very little empirical data available to actually determine which is most common. Moreover, regardless of the specifics for this particular correlation, the general pattern and considerations presented here will apply in similar ways to correlations between other aspects of the acute stress response.

@@

The time that samples were taken at, relative amount of variation in speed and maximum, and degree of correlation between the speed and maximum all had substantial impacts on the ability to infer true trait values from single time point glucocorticoid measures (figure \@ref(fig:cor-ex2)). While these scenarios do not explore all possible parameter space, there are several clear conclusions that can be made. First, neither speed or maximum traits could be assessed accurately when between-individual variation in both traits was low (figure \@ref(fig:cor-ex2)A). This is potentially important for interpreting apparent differences in glucocorticoid fitness relationships because between-individual variation is known to differ across life history stages [CITE]. Second, accurately assessing variation in speed was much harder-if not impossible-with single measures.

Only when speed varied between individuals with very low variation in maximum was it possible to accurately measure speed, but this situation may be rare in natural populations. When speed was tightly correlated with maximum (figure \@ref(fig:cor-ex2)D) it was sometimes possible to attain reasonable estimates of speed (figure \@ref(fig:cor-ex2)C-D), but when speed was not correlated with maximum single measures were not good indicators of variation in speed (figure \@ref(fig:cor-ex2)A, C-D). Finally, measuring variation in maximum values was much easier under many conditions (figure \@ref(fig:cor-ex2)C-D), but the accuracy of assessment of maximum values was also negative impacted by variation in speed and the degree of this impact differed depending on the correlation between the two traits (figure \@ref(fig:cor-ex2)). Beyond the specifics of this particular example, what these results demonstrate clearly is that understanding what aspect of the glucocorticoid response is being measured by any particular study design depends on extensive knowledge of the overall shape and amount of variation in different aspects of the acute stress response.

### *Detecting links between fitness and responses*

A common goal of recent studies is to establish whether variation in glucocorticoids is associated with fitness or some proxy for fitness [@schoenle2020]. While there has been a great deal of discussion about the extent to which these relationships might differ with life history characteristics or between breeding stages [CITE], there has been relatively little consideration of the way that methodological limitations might limit the ability to detect these relationships even when they exist.

@@

Several patterns can be identified by examining the results of this simulation. First, the correlation between the true maximum glucocorticoid value and fitness does not differ for populations simulated with high or low between-individual variation (figure \@ref(fig:fitness-ex) A-D). In all cases, however, the observed correlation is always lower than the true correlation and always lowest in the population with low between-individual variation. The ubiquity of this pattern is a product of the simulation structure, because adding measurement error or within-individual variation effectively adds noise to the true correlation. It is important to note that in the real world, it is unlikely that this pattern would be so universal, because unmeasured variables could influence both fitness and glucocorticoids. For example, if habitat quality directly alters fitness and glucocorticoids, the observed correlation could be stronger than the 'true' correlation. Thus, interpretation of these results should be made cautiosly in light of the simplicity of the simulation compared to real world conditions.

Nevertheless, general patterns illustrated by the simulation are likely to pertain across a wide range of conditions. In this case, it is easiest to detect significant glucocorticoid-fitness relationships when both measurement error and within-individual variation are low (figure \@ref(fig:fitness-ex) A). It becomes harder to detect these true relationships when either measurement error (figure \@ref(fig:fitness-ex) B) or within-individual variation (figure \@ref(fig:fitness-ex) C) are high, but even in these more challenging situations the relationship can be detected the majority of the time if between-individual variation in maximum levels is high. When both measurement error and within-individual variation are high, it is nearly impossible to detect glucocorticoid-fitness relationships with low-between individual variation, but in populations with high between-individual variation the relationship is still detected in about half of the simulations.

The fact that low between-individual variation in maximum glucocorticoids makes it harder to detect true glucocorticoid-fitness relationships across a wide range of conditions has important consequences for interpreting empirical results. Many studies have demonstrated different relationships (or lack thereof) between corticosterone and fitness at different life history stages [add our FEE & Bonier papers CITE], but it is also well known that the absolute amount of between individual variation in glucocorticoid traits varies considerably at different stages [CITE]. Our simulation demonstrates that the power to detect true relationships will differ drastically across these conditions even with similar study designs and samples sizes, suggesting that great care is needed to conclusively demonstrate different glucocorticoid-fitness relationships across contexts.

### *Designing optimal sampling strategies*

One of the major benefits of simulating glucocorticoid response curves will be the ability to design optimal sampling strategies before data are collected. A simulation can be constrained to match any real world limitations (e.g., maximum number of samples possible per individual) and then explored to determine how to best allocate sampling resources. The specifics of this task will vary considerably with the study system and question being addressed, but here I illustrate one possible application. Consider an experiment in which the acute glucocorticoid response of a treatment group and control group are compared after some experimental manipulation. The details of the manipulation are unimportant here, but suppose that the prediction is that this manipulation should result in a difference in the speed of the corticosterone response between our two groups, such that the treatment group will reach it's maximum glucocorticoid value faster than the control group, but will not differ in the maximum value itself. I have implemented this difference by simulating two populations in which the treatment group has a steeper initial slope and also reaches the maximum value faster (figure \@ref(fig:prediction-ex)).

@@

Next, we can ask how well different study designs can detect this difference. Here we can impose any logistical constraints relevant to the study system. As an example, in this case we can only sample a maximum of 20 individuals per group, we can only sample each individual once post-treatment, and during that single sampling event we can take a blood sample at a maximum of two different time points, resulting in a total of 80 data points. Given these constraints, I compare three different sampling designs: i) a study in which every animal is sampled at 1 minute, 30 minutes, and 60 minutes, ii) a study in which two sampling times between 1 and 60 minutes are randomly chosen for every animal, iii) a study in which two sampling times are randomly chosen for each animal, but weighted more heavily around the range of times when maximum levels are expected to be reached for the population. Note that the first sampling scheme closely mirrors the most common empirical design and in this case I have allowed an extra, third sample at 60 minutes, such that it includes 120, rather than 80, data points. For illustration purposes I sampled directly from the 'true' response curves in this example so that there is no additional measurement error added. To evaluate these schemes I compare estimates of the acute response curve for each group to the 'true' known curves shown in figure \@ref(fig:prediction-ex). Note that a more complete analysis of a sampling schemes performance should include many more iterations and full statistical comparisons, but the details here will be highly dependent on the study system and goals, so I provide this simple example to illustrate the approach rather than to make any more widely applicable conclusions.

@@

In this case, the standard sampling scheme performs poorly (figure \@ref(fig:treatment-ex) A), with no differences detectable between the two groups, despite the fact that the treatment group reaches it's maximum value on average 12 minutes (~40%) faster than the control group. In contrast, both the random sampling and weighted sampling schemes detect differences in the shape of the acute response (figure \@ref(fig:treatment-ex) B & C). In this particular scenario, there is no clear difference between these two approaches. A few clear takeaways can be derived from these results. First, while strict standardization of the timing of samples has some clear advantages, it also comes with costs and likely makes it nearly impossible to detect certain types of variation between groups or individuals. In this case, standardized sampling performed much worse than the other two approaches despite the fact that the analysis included 50% more data and it should be clear that no amount of additional sampling would allow that approach to detect the differences between these groups. Second, while it may be very difficult to accurately estimate the full shape of the acute stress response for *individuals*, the sampling schemes shown here demonstrate that it should be possible to describe these shapes accurately for groups (e.g., treatments, species, different contexts) even without extraordinarily large sample sizes. A similar argument about the power of randomly timed sampling has been put forward in the function valued trait literature [CITE], but this type of sampling scheme is rarely used in evolutionary endocrinology research.

This simulation is particular to a single very specific scenario, but a similar scenario could be designed for any number of studies and any number of predictions about how the speed, scope, or other attributes of the glucocorticoid response are expected to change with a treatment or between different groups or species. Clearly, when estimating the timing of peak glucocorticoids, a simple baseline plus induced sampling scheme is sub optimal, but this scheme may be perform well in other situations where the maximum value is the target and there is relatively little variation in response time. Creating simulations like this before studies are conducted has the potential increase the efficient use of researches time and funds, but also forces researches to think explicitly about quantitative predictions ahead of time. These simulations could be included as part of a study pre-registration, grant application, or registered report to demonstrate exactly what data collection and analysis approaches are planned and to justify those decisions.

## CONCLUSION

>>>>

---
title: "Simulating physiological flexibility in the acute glucocorticoid response to stressors reveals limitations of current empirical approaches"
author: 
 - Conor C. Taff
header-includes:
  - \usepackage[font={footnotesize}, labelfont={bf}]{caption}
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}\linenumbers
fontsize: 12pt
output: 
  word_document:
  bookdown::pdf_document2:
      number_sections: FALSE
      toc: FALSE
      extra_dependencies: ["flafter"]
  bookdown::word_document2:
      number_sections: FALSE
      toc: FALSE
      extra_dependcies: ["flafter"]
bibliography: references.bib
csl: animal-behaviour.csl
---

@@

*Cornell Lab of Ornithology and Department of Ecology & Evolution*

*cct63@cornell.edu*

\newpage
\raggedright

@@

Wild animals often experience unpredictable challenges that demand rapid and flexible responses. The glucocorticoid mediated stress response is one of the major systems that allows vertebrates to rapidly adjust their physiology and behavior. Given its role in responding to challenges, evolutionary physiologists have focused on the consequences of between-individual and, more recently, within-individual variation in the acute glucocorticoid response. However, empirical studies of physiological flexibility are severely limited by the logistical challenges of measuring the same animal multiple times. Data simulation is a powerful approach when empirical data are limited, but has not been adopted to date in studies of physiological flexibility. In this paper, I develop a simulation that can generate realistic acute glucocorticoid response data with user specified characteristics. Simulated animals can be sampled continuously through an acute response and across as many separate responses as desired, while varying key parameters. Using the simulation, I develop several scenarios that address key questions in physiological flexibility. These scenarios demonstrate the conditions under which a single glucocorticoid trait can be accurately assessed with typical experimental designs, the consequences of covariation between different components of the acute stress response, and the way that context specific differences in variability of acute responses can influence the power to detect relationships between the strength of the acute stress response and fitness. I also describe how to use the simulation tools to aid in the design and evaluation of empirical studies. Recently there has been a great deal of interest in understanding the causes and consequences of physiological flexibility, but empirical data are often extremely limited and traditional sampling methods are poorly designed to directly address the questions of interest. This simulation represents a way forward by revealing critical aspects of physiological flexibility and by creating a tool for designing better empirical studies that integrate simulation and theory with data collection.

*Keywords: acute stress response, physiological flexibility, glucocorticoids, evolutionary endocrinology*

@@

The recognition that characterizing the functional shape of an acute stress response is challenging goes back to the earliest studies conducted in wild animals. Early studies often employed various control groups and sampled individual animals at a variety of time points over a long period in order to describe the full response curve for a particular group [e.g., a species or a breeding stage, @wingfield1992]. These validations were considered essential to characterize key parameters of the acute response for each group being studied (i.e., baseline, rate of increase, maximum level, time of peak, and area under the curve; John Wingfield, *personal communication*). The challenge of estimating these parameters becomes much more difficult when trying to describe the response for an individual animal rather than for a group, because glucocorticoids can often only be measured at two or three time points and only a small number of times per animal [e.g., @vitousek2018]. Because these studies require an estimate for each individual, the solutions used by older studies that added additional animals to allow for sampling at more time points are not available.

@@

One of the characteristics of both the within-individual reaction norm and FVT literature is that empirical work has proceeded in very close coordination with simulation and statistical method development. In contrast, studies of endocrine flexibility often point to these methods, but don't address the ways that the particular logistical challenges of hormone measurement might necessitate different empirical approaches. I believe this is one reason that there are currently more conceptual papers arguing for a reaction norm approach to endocrine variation than there are empirical papers actually applying the approach. While many of the tools developed in these related fields are transferable, studies of physiological flexibility would benefit from a focus on analysis development and testing that explicitly incorporates the particular details and challenges of these questions. One way to accomplish these goals is to use simulations, but to my knowledge no studies of physiological flexibility have developed simulations of the acute stress response that address the issues discussed above.

@@

## MATERIALS AND METHODS

### DESCRIPTION OF THE SIMULATION

I developed a set of functions in R version 4.0.2 [@rcore] to generate acute physiological response curves. This simulation makes no assumptions about the  mechanistic process that results in the shape of a glucocorticoid response. Rather, parameters are sampled to generate curves that are similar in shape and degree of variation to empirically observed responses (Figure \@ref(fig:concept-fig)). This simulation is designed to create data sets with realistic structure that can be used to better design and plan studies of physiological flexibility, to evaluate power of current study designs, and to evaluate the sensitivity of sampling regimes to any number of modifications to the shape of glucocorticoid response curves (e.g., changing covariation patterns between different features of the response). I explore a small number of scenarios in the next section, but I expect that many other scenarios can be addressed with these tools. For illustration purposes, I refer to simulated glucocorticoid responses, but the simulation applies equally well to any physiological mediator of a rapid response. The package can be installed in R using the following command.

@@

A second function, `cort_sim2`, starts with a population of animals generated from `cort_sim1` and samples observed acute glucocorticoid responses an arbitrary number of times for each animal. Two sources of variation in the observed relative to true parameter values can be specified. First, within-individual variation in expression is represented by specifying what amount of variation in the observation of each parameter is determined by the true value and what amount is determined by an additional randomly sampled response, based on the population parameters (this additional sampling maintains the user specified covariance structure of the population). After sampling the parameters, values are interpolated for each one minute time point and a localized regression is fit to create a smoothed curve that represents the observed glucocorticoid response. From this expressed response, individual data points are then collected at user specified times that would reflect an empirical study design (e.g., 2, 30, and 60 minutes). Additional noise can be added to these data points to represent measurement error (e.g., assay error).

@@

Finally, given recent interest in estimating the repeatability of glucocorticoid regulation [@taff2018; @cockrem2013; @hau2016], I also included a function that takes input from `cort_sim2` and calculates the observed repeatability of several measures using package `rptR` [@stoffel2017]. Full details are included in the package documentation, but this function returns repeatability for each individual time point specified in the down sampled data set, profile repeatability [@reed2019], and repeatability for area under the curve calculated as both increase (AUC~I~) and ground (AUC~G~) approaches [@pruessner2003]. For each AUC measure, the function returns repeatability for the full time course, for an estimate using only the observed values in the down sampled data set, and for the full data set constrained to the time period encompassing the observed data points. Simple plots illustrating repeated samples from the same individuals are also returned by default. I do not develop an example of repeatability in this manuscript, but the functions here could be used to determine the impact of different study design choices on repeatability estimates.

### RESULTS

@@

#### *Simulating empirically parameterized data*

@@

#### *Accurately measuring a single glucocorticoid trait*

Single time point measures of glucocorticoids are often interpreted as representing meaningful variation between individuals. For example, variation in the level of glucocorticoids after 30 minutes of standardized restraint is typically interpreted as variation in the magnitude of the stress response [@taff2019]. However, this interpretation rests on assumptions that are rarely explicitly tested with empirical data. For example, the time chosen to take a stress-induced sample is often assumed to be either at the species peak or during a plateau period after the species peak. In some early studies, great care was taken to determine an average population level peak time [@wingfield1992], but many studies adopt the widely used 'standard' time of 30 minutes post capture without extensive validation [compiled in @vitousek2019]. While there is a general assumption that sampling later than the peak is acceptable (and perhaps preferable) because animals will be sampled during a relatively stable high plateau, there is little empirical data to evaluate this assertion or to determine how much under or overshooting the species peak timing might influence inferences. Furthermore, even when the average peak timing is well established, differences in the amount of between-individual variation in the time to reach the peak or in peak values are common across species and even in different life history stages within species [@wingfield1992]. The combinations of these patterns of variation could have major consequences on the accuracy of single point estimates taken at 30 minutes, but these questions cannot be addressed directly with empirical datasets where the true underlying values of each individual are unknown.

@@

Results of this simulation are summarized in figure \@ref(fig:max-sim). The amount of between-individual variation in the maximum glucocorticoid value has a profound effect on the ability to detect true maximal levels with samples taken at 30 minutes. In one sense, this result is unsurprising because it is intuitive that large differences would be easier to detect, but there are important consequences of this fact for interpreting studies that seek to link between-individual variation in the magnitude of the stress response with other traits. For example, the magnitude of the acute stress response often varies substantially across life history stages [@wingfield1992]. Even if study designs are identical and maximum glucocorticoids are associated with performance, it will be easier to detect those patterns during life history stages with greater variation (see section on detecting fitness associations below). There is a weaker, but still substantial impact of variation in the time taken to reach maximum values on the accuracy of estimates in this simulation. Greater variation in the speed of the response reduces the accuracy of estimates of maximal values. Finally, the timing of sampling relative to the average population peak timing also influences accuracy. Measuring after the average peak time results in the most accurate estimates across a range of parameter values, while measuring before the average peak time produces the least accurate measures, particularly when there is also high variation in the time to reach maximum values between individuals. This simple example demonstrates clearly that the same experimental design will perform better or worse depending on the combination of glucocorticoid regulation parameters in the population being studied.

#### *Exploring covariance between response components*

In reality, fully characterizing the acute glucocorticoid response requires more than identifying just the maximum value reached. Individuals may differ in baseline levels, rate of initial increase, the speed of reaching the maximum level, time spent at maximum, and the speed of return to baseline. Moreover, each of these components of the endocrine response could be positively or negatively correlated with each other within and between individuals. In these cases, measurements taken at particular time points contain information about multiple aspects of the response and without additional information it may be difficult to know what trait is being measured. The fact that each of these traits might be important and that they might covary has been discussed in a general sense [e.g., @baugh2013], but simulations are uniquely powerful for exploring under exactly what conditions time point measure of glucocorticoids can or cannot be used as indicators of these traits.

@@

To illustrate this point, I explored the consequences of variation in the correlation between and relative amount of variation in just two aspects of the acute stress response: the maximum glucocorticoid level reached and the time required to reach the maximum level. For simplicity, I refer to the 'speed' of the response, but note that other aspects, such as the rate of initial increase, could also be considered as variation in the speed of response. When considering these two traits, a population of animals could plausibly display one of three patterns. Individuals that reach their maximum value faster might also reach higher values (figure \@ref(fig:cor-ex1)A; simulation correlation = -0.6). Alternatively, the speed and maximum values might vary independently (figure \@ref(fig:cor-ex1)B; correlation = 0). Finally, individuals that are faster responders might max out at lower glucocorticoid values (figure \@ref(fig:cor-ex1)C; correlation = 0.6). While many researchers in this field might have intuitions about which of these scenarios is most likely to prevail, there is very little empirical data available to actually determine which is most common. Moreover, regardless of the specifics for this particular correlation, the general pattern and considerations presented here will apply in similar ways to correlations between other aspects of the acute stress response.

@@

The time that samples were taken at, relative amount of variation in speed and maximum, and degree of correlation between the speed and maximum all had substantial impacts on the ability to infer true trait values from single time point glucocorticoid measures (figure \@ref(fig:cor-ex2)). While these scenarios do not explore all possible parameter space, there are several clear conclusions that can be made. First, neither speed or maximum traits could be assessed accurately when between-individual variation in both traits was low (figure \@ref(fig:cor-ex2)A). This is potentially important for interpreting apparent differences in glucocorticoid fitness relationships because between-individual variation is known to differ across life history stages [@wingfield1992]. Second, accurately assessing variation in speed was much harder---if not impossible---with single measures.

It was only possible to accurately estimate speed when high between-individual variation in speed was coupled with low variation in maximal values, but this situation may be rare in natural populations. When speed was tightly correlated with maximum (figure \@ref(fig:cor-ex2)D) it was sometimes possible to attain reasonable estimates of speed (figure \@ref(fig:cor-ex2)C-D), but when speed was not correlated with maximum, single measures were not good indicators of variation in speed (figure \@ref(fig:cor-ex2)A, C-D). Finally, measuring variation in maximum values was much easier under many conditions (figure \@ref(fig:cor-ex2)C-D), but the accuracy of assessment of maximum values was also negative impacted by variation in speed and the degree of this impact differed depending on the correlation between the two traits (figure \@ref(fig:cor-ex2)). Beyond the specifics of this particular example, what these results demonstrate clearly is that understanding what aspect of the glucocorticoid response is being measured by any particular study design depends on extensive knowledge of the overall shape and amount of variation in different aspects of the acute stress response.

#### *Detecting links between fitness and responses*

A common goal of recent studies is to establish whether variation in glucocorticoids is associated with fitness or some proxy for fitness [@schoenle2020]. While there has been a great deal of discussion about the extent to which these relationships might differ with life history characteristics or between breeding stages, there has been relatively little consideration of the way that methodological limitations might limit the ability to detect these relationships even when they exist.

@@

Several patterns can be identified by examining the results of this simulation. First, the correlation between the true maximum glucocorticoid value and fitness does not differ for populations simulated with high or low between-individual variation (figure \@ref(fig:fitness-ex)A-D). In all cases, however, the observed correlation is lower than the true correlation and always lowest in the population with low between-individual variation. The ubiquity of this pattern is a product of the simulation structure, because adding measurement error or within-individual variation effectively adds noise to the true correlation. It is important to note that in the real world, it is unlikely that this pattern would be so universal, because unmeasured variables could influence both fitness and glucocorticoids. For example, if habitat quality directly alters fitness and glucocorticoids, the observed correlation could be stronger than the 'true' correlation. Thus, interpretation of these results should be made cautiosly in light of the simplicity of the simulation compared to real world conditions.

Nevertheless, general patterns illustrated by the simulation are likely to pertain across a wide range of conditions. In this case, it is easiest to detect significant glucocorticoid-fitness relationships when both measurement error and within-individual variation are low (figure \@ref(fig:fitness-ex)A). It becomes harder to detect these true relationships when either measurement error (figure \@ref(fig:fitness-ex)B) or within-individual variation (figure \@ref(fig:fitness-ex)C) are high, but even in these more challenging situations the relationship can be detected the majority of the time if between-individual variation in maximum levels is high. When both measurement error and within-individual variation are high, it is nearly impossible to detect glucocorticoid-fitness relationships with low-between individual variation, but in populations with high between-individual variation the relationship is still detected in about half of the simulations.

The fact that low between-individual variation in maximum glucocorticoids makes it harder to detect true glucocorticoid-fitness relationships across a wide range of conditions has important consequences for interpreting empirical results. Many studies have demonstrated different relationships (or lack thereof) between corticosterone and fitness at different life history stages [@vitousek2018; @bonier2009b], but it is also well known that the absolute amount of between individual variation in glucocorticoid traits varies considerably at different stages [@wingfield1992]. Our simulation demonstrates that the power to detect true relationships will differ drastically across these conditions even with identical study designs and samples sizes, suggesting that great care is needed to conclusively differentiate true differences in glucocorticoid-fitness relationships across contexts from statistical artefacts.

#### *Designing optimal sampling strategies*

One of the major benefits of simulating glucocorticoid response curves will be the ability to design optimal sampling strategies before data are collected. A simulation can be constrained to match any real world limitations (e.g., maximum number of samples possible per individual) and then explored to determine how to best allocate sampling resources. The specifics of this task will vary considerably with the study system and question being addressed, but here I illustrate one possible application. Consider an experiment in which the acute glucocorticoid response of a treatment group and control group are compared after some experimental manipulation. The details of the manipulation are unimportant here, but suppose that the prediction is that this manipulation should result in a difference in the speed of the corticosterone response between our two groups, such that the treatment group will reach it's maximum glucocorticoid value faster than the control group, but will not differ in the maximum value itself. I have implemented this difference by simulating two populations in which the treatment group has a steeper initial slope and also reaches the maximum value faster (figure \@ref(fig:prediction-ex)). Any number of possible hypotheses for a particular study system could be specified following a similar approach.

@@

Next, we can ask how well different study designs can detect this difference. Here we can impose any logistical constraints relevant to the study system. As an example, in this case we can only sample a maximum of 20 individuals per group, we can only sample each individual once post-treatment, and during that single sampling event we can take a blood sample at a maximum of two different time points, resulting in a total of 80 data points. Given these constraints, I compare three different sampling designs: i) a study in which every animal is sampled at 1 minute, 30 minutes, and 60 minutes, ii) a study in which two sampling times between 1 and 60 minutes are randomly chosen for every animal, iii) a study in which two sampling times are randomly chosen for each animal, but weighted more heavily around the range of times when maximum levels are expected to be reached for the population.

Note that the first sampling scheme closely mirrors the most common empirical design and in this case I have allowed an extra, third sample at 60 minutes, such that it includes 120, rather than 80, data points. For illustration purposes I sampled directly from the 'true' response curves in this example so that there is no additional measurement error added. To evaluate these schemes I compare estimates of the acute response curve for each group to the 'true' known curves shown in figure \@ref(fig:prediction-ex). Note that a more complete analysis of a sampling schemes performance should include many more iterations and full statistical comparisons, but the details here will be highly dependent on the study system and goals, so I provide this simple example to illustrate the approach rather than to make any more widely applicable conclusions.

@@

In this case, the standard sampling scheme performs very poorly (figure \@ref(fig:treatment-ex)A), with no differences detectable between the two groups, despite the fact that the treatment group reaches it's maximum value on average 12 minutes (~40%) faster than the control group. In contrast, both the random sampling and weighted sampling schemes detect differences in the shape of the acute response (figure \@ref(fig:treatment-ex) B & C). In this particular scenario, there is no clear difference between these two approaches. A few clear takeaways can be derived from these results. First, while strict standardization of the timing of samples has some clear advantages, it also comes with costs and likely makes it nearly impossible to detect certain types of variation between groups or individuals. In this case, standardized sampling performed much worse than the other two approaches despite the fact that the analysis included 50% more data; it should be clear that no amount of additional sampling would allow that approach to detect this particular pattern of between group differences. Second, while it may be very difficult to accurately estimate the full shape of the acute stress response for *individuals*, the sampling schemes shown here demonstrate that it should be possible to describe these shapes accurately for groups (e.g., treatments, species, different contexts) even without extraordinarily large sample sizes. A similar argument about the power of randomly timed sampling has been put forward in the function valued trait literature [@gomulkiewicz2018], but this type of sampling scheme is rarely used in evolutionary endocrinology research. It is perhaps unsurprising that the few empirical papers that have emphasized the importance of different time courses (rather than only maximum) of the stress response have often focused on between group comparisons or investigated variation in the exact sampling time between individuals [e.g., @baugh2013; @small2017]
